Left-linear rewriting in adhesive categories

### 2024-04-23 concur/fusione.tex
minor
minori

\newcommand{\gph}[1]{\mathcal{G}\textbf{\textup{-Graph}}}

@@

    \caption{Rules of the rewriting system of the running example.}
    \label{fi:rules}
  \end{figure}

@@

\begin{figure}
  \begin{subfigure}{0.2\textwidth}
    {\small
      \xymatrix@C=16pt@R=1pt{
        & 0 \ar@{-}[dd] \ar@{-}[ddddddl] \ar@{-}[ddddddr] &  \\
        &  & \\
        & 1 \ar@{-}[dd] \ar@{-}[ddddl]  \ar@{-}[ddddr]   &  \\
        &  & \\
        & 2 \ar@{.}[d] \ar@{-}[ddl]   \ar@{-}[ddr]     &  \\
        &  \ar@{.}[dl]   \ar@{.}[dr] & \\
        b \ar@{-}[dr] & & c \ar@{-}[dl]\\
        & a 
      }
    }
    \caption{Poset $(P, \leq)$}
    \label{fig:diff1:P}
  \end{subfigure}
  \begin{subfigure}{0.58\textwidth}
    \xymatrix@C=12pt{
      a \ar[d]_{m_0} &
      a \ar[d]_{k_0} \ar[l]_{\id{a}} \ar[rr]^{(a,0)} & &
      0 \ar@/^.35cm/[drrrr]|(.315)\hole^(.75){\id{0}}
      \ar[dr]|(.3)\hole_{\id{0}} & & 
      a \ar@/_.35cm/[dllll]_(.75){(a,c)}
      \ar[dl]|(.3)\hole^{(a,0)} & &
      a \ar[d]^{(a,0)}\ar@{>->}[ll]_{\id{a}} \ar[rr]^{(a,b)} &&
      b \ar[d]^{(b,0)}\\
      c &
      \ar@{>->}[l]^{\id{c}} c \ar[rrr]_{(c,0)}& & & 0 &&&
      \ar@{>->}[lll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0
    }
    \caption{Two rewriting steps}
    \label{fig:diff1:rew}
  \end{subfigure}
  \begin{subfigure}{0.20\textwidth}
    \xymatrix@C=16pt{a \ar[d]_{(a,c)}& a
      \ar[d]_{(a,c)}\ar@{>->}[l]_{\id{a}} \ar[r]^{(a,b)} & b \\c & c
      \ar[l]^{\id{c}}
    }
    \caption{No rewriting step}
    \label{ex:diff1:no-step}
  \end{subfigure}
  \caption{Sequential independence does not imply switchability}
  \label{fig:diff1} 
\end{figure}

\begin{example}
  \label{ex:diff1}
  Consider the poset $(P, \sqsubseteq)$ in Fig.~\ref{fig:diff1:P} where
  $P = \mathbb{N} \cup \{a,b,c\}$ and $\sqsubseteq$ is defined by
  $m \sqsubseteq n$ if $m \geq n$, $a \sqsubseteq x$ for all $x \in P$
  and $b, c \sqsubseteq n$ for all $n \in \mathbb{N}$.
  %
  Let $\X$ be the  
  category associated with such order, which by \Cref{rem:iso} is
  $\mathsf{Iso(\X)}$-adhesive. 
  %
  Consider a rewriting
  system whose set of rules contains the following 
  \[\xymatrix{a & a \ar[r]^{(a, 0)} \ar[l]_{\id{a}} & 0 & a & a
      \ar[r]^{(a, b)} \ar[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ in Fig.~\ref{fig:diff1:rew}.
  % \[
  %   \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
  %     \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
  %     \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
  %     \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
  %     \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
  %     \ar@{>->}[ll]^{\id{c}} c \ar[rr]_{(c,0)}&& 0 &&
  %     \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]
  %
  Note that the two direct derivations are sequential
  independent. However there is no switch since the rule applied by
  $\dder{D}_1$ cannot be applied to $c$. In fact, the unique morphism
  $a\to c$ is $(a,c)$, thus we get the diagram in Fig.~\ref{ex:diff1:no-step}. But $b$ and $c$
  do not have a supremum in the poset underlying $\X$, thus the arrows
  $(a,b)\colon a\to b$ and $(a,c)\colon a\to c$ do not have a
  pushout. Hence we do not get a direct derivation from $c$.
\end{example}

@@

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category
  which is not strong enforcing. The example is somehow
  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but it needs to be so, since natural examples are all in the well-beahved class mentioned above. 
  .\todo{P: Scritto qualcosa, probabilmente da migliorare}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

A first class consists of those rewriting systems over
possibly hierarchical graphical structures obtained as algebras of
suitable signatures where rules are constrained not to merge elements
of top level sorts in the hierarchy (e.g., for graphs, nodes can be
merged while edges cannot). We first introduce signatures that are
essentially acyclic graphs whose nodes can be seen as sorts.

\begin{definition}
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph where $s,t\colon E\rightrightarrows V$.
  We define the \emph{depth} of a vertex $v\in V$ as\todo{ma ci serve tutta la depth?}
  \[
    \dph(v)=\begin{cases}
      0 & \text{there is no $e\in E$ such that $t(e)=v$}\\
      m+1 &\text{where } m=\min\{\dph(s(e)) \mid e\in E \text{ such that } t(e)=v
      \} 
    \end{cases}
  \]
\end{definition}

@@

\begin{definition}[Node-merging graphical rewriting systems]
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph
  and $\G$ the free category over $\mathcal{G}$. The
  category $\gph{G}$ of \emph{$\mathcal{G}$-graphs} is the category
  $\Set^{\G}$. A
  \emph{node-merging graphical rewriting system}\todo{altro nome? source/leaf preserving} is a left-linear
  rewriting system $(\gph{G}, \R)$ such that  the
  following hold for each rule
  $(l\colon K\to L, r\colon K\to R)$ in $\R$
  \begin{enumerate}
  \item for every $v\in V$ and $x\in L(v)$, there exists a vertex $w$
    with depth $0$ and an arrow $p\colon w\to v$ (i.e.~a path in
    $\mathcal{G}$) such that $x$ belongs to the image of
    $L(p)\colon L(w)\to L(v)$;
  \item $r\colon K\to R$ is mono on vertices of depth $0$, i.e.~for every
    $v\in V$ such that $\dph(v)=0$, the component $r_v:K(v)\to R(v)$
    is mono.
  \end{enumerate}
\end{definition}

For instance, the category $\cat{Graph}$ can be obtained
by taking as $\mathcal{G}$ the graph $e \rightrightarrows v$. The sort
of edges $e$ has depth $0$ while $v$, the sort of nodes, has depth
$1$. Hence, condition $1$ asks that in the left-hand side of each rule
there are no isolated nodes, while condition $2$ asks that the
morphism $r: K \to R$ is mono on edges, i.e.~it can only merge nodes.

@@

\begin{restatable}{lemma}{lemVTame}
\label{bono}
  All node-merging graphical rewriting systems are well-switching.
\end{restatable}

@@

  \begin{figure}[t]
    \input{Figure/rules-cas.tex}    
    \caption{Additional rules.}
    \label{fi:rules-cas}
  \end{figure}

@@

\begin{remark}\label{rem:po}
	Notice that $\gph{G}$ is adhesive and that pushouts are computed component-wise, thus if $g\colon F\to G$ is the pushout of $r\colon K\to R$, then $g_v\colon F(v)\to G(v)$ is mono for every $v$ with depth $0$.
\end{remark}

@@

\begin{proof}
	Every node-merging graphical rewriting system is strong enforcing because $\gph{G}$ is in the class $\mathbb{B}^+$ defined in \Cref{app:fill}.  Consider the following diagram 
	\[\xymatrix@R=25pt@C=30pt{L_0 \ar[d]_{m_0}&& K_0
		\ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
		\ar@<-.5ex>@/^.35cm/[drrr]|(.23)\hole|(.34)\hole_(.4){i_0} 	\ar@<.5ex>@/^.35cm/[drrr]^(.2){j_0}|(.31)\hole |(.4)\hole
		\ar[dr]|(.3)\hole_{h_0} && L_1 \ar@<.5ex>@/_.35cm/[dlll]^(.4){i_1} \ar@<-.5ex>@/_.35cm/[dlll]_(.2){j_1}
		\ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
		\ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
		D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
		\ar[rr]_{g_1}&& G_2}\]

	Since $f_1$ is mono then $i_0=j_0$. For every $v$ with depth $0$ we have
	\[g_{0,v}\circ i_{1,v}=m_{1,v}=g_{0,v}\circ j_{1,v}\]
	which entails $i_{1,v}=j_{1,v}$. Let now $v$ be a vertex with depth greater than $0$, and take $x\in L_1(v)$. By definition, we have $p:w\to v$ and $y\in L_{1}(w)$  such that $\dph(w)=0$ and $L(p)(y)=x$. By naturality we have
	\[i_{1,v}(x)=i_{1,v}(L(p)(y))=L(p)(i_{1, w}(y))=L(p)(j_{1, w}(y))=j_{1,v}(L(p)(y))=j_{1,v}(x)\]
	From this we can deduce that $i_{1,v}=j_{1,v}$, as wanted.
\end{proof}

>>>>

\newcommand{\gph}[1]{\textbf{\textup{{#1}-Graph}}}

@@

    \caption{A rewriting system in $\cat{Graph}$.}
    \label{fi:rules}
  \end{figure}

@@

\begin{figure}
  \begin{subfigure}{0.2\textwidth}
    {\small
      \xymatrix@C=16pt@R=1pt{
        & 0 \ar@{-}[dd] \ar@{-}[ddddddl] \ar@{-}[ddddddr] &  \\
        &  & \\
        & 1 \ar@{-}[dd] \ar@{-}[ddddl]  \ar@{-}[ddddr]   &  \\
        &  & \\
        & 2 \ar@{.}[d] \ar@{-}[ddl]   \ar@{-}[ddr]     &  \\
        &  \ar@{.}[dl]   \ar@{.}[dr] & \\
        b \ar@{-}[dr] & & c \ar@{-}[dl]\\
        & a 
      }
    }
    \caption{Poset $(P, \sqsubseteq)$}
    \label{fig:diff1:P}
  \end{subfigure}
  \begin{subfigure}{0.58\textwidth}
    \xymatrix@C=20pt{
      a \ar[d]_{\sqsubseteq} & a \ar[d]_{\sqsubseteq} \ar@{>->}[l]_{\id{a}} \ar[r]^{\sqsubseteq} & 
      0 \ar@{>->}@/^.35cm/[drrr]|(.3)\hole^(.75){\id{0}}
      \ar@{>->}[dr]|(.3)\hole_{\id{0}} &  &
      a \ar@/_.35cm/[dlll]_(.75){\sqsubseteq}
      \ar[dl]|(.3)\hole^{\sqsubseteq} & 
      a \ar[d]^{\sqsubseteq}\ar@{>->}[l]_{\id{a}} \ar[r]^{\sqsubseteq} &
      b \ar[d]_{\sqsubseteq}\\
      c &
      \ar@{>->}[l]^{\id{c}} c \ar[rr]_{\sqsubseteq}& &  0 &&
      \ar@{>->}[ll]^{\id{0}} 0 \ar@{>->}[r]_{\id{0}}& 0
    }
    \caption{Two rewriting steps}
    \label{fig:diff1:rew}
  \end{subfigure}
  \begin{subfigure}{0.20\textwidth}
    \xymatrix@C=20pt{a \ar[d]^{\sqsubseteq}& a
      \ar[d]^{\sqsubseteq}\ar@{>->}[l]_{\id{a}} \ar[r]^{\sqsubseteq} & b \\c & c
      \ar@{>->}[l]^{\id{c}}
    }
    \caption{No rewriting step}
    \label{ex:diff1:no-step}
  \end{subfigure}
  \caption{Sequential independence does not imply switchability}
  \label{fig:diff1} 
\end{figure}

\begin{example}
  \label{ex:diff1}
  Consider the poset $(P, \sqsubseteq)$ in Fig.~\ref{fig:diff1:P} where
  $P = \mathbb{N} \cup \{a,b,c\}$ and $\sqsubseteq$ is defined by
  $m \sqsubseteq n$ if $m \geq n$, $a \sqsubseteq x$ for all $x \in P$
  and $b, c \sqsubseteq n$ for all $n \in \mathbb{N}$.
  %
  Let $\X$ be the  
  category associated with such order, which by \Cref{rem:iso} is
  $\mathsf{Iso(\X)}$-adhesive. 
  %
  Consider a rewriting
  system whose set of rules contains the following 
  \[\xymatrix{a & a \ar[r]^{\sqsubseteq} \ar@{>->}[l]_{\id{a}} & 0 & a & a
      \ar[r]^{\sqsubseteq} \ar@{>->}[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ in Fig.~\ref{fig:diff1:rew}.
  % \[
  %   \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
  %     \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
  %     \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
  %     \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
  %     \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
  %     \ar@{>->}[ll]^{\id{c}} c \ar[rr]_{(c,0)}&& 0 &&
  %     \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]
  %
  Note that the two direct derivations are sequential
  independent. However there is no switch since the rule applied by
  $\dder{D}_1$ cannot be applied to $c$. In fact, there is a unique morphism
  $a\to c$, yielding the diagram in Fig.~\ref{ex:diff1:no-step}. But $b$ and $c$
  do not have a supremum in the poset underlying $\X$, thus the arrows
  $a\to b$ and $a\to c$ do not have a
  pushout. Hence we do not get a direct derivation from $c$.
\end{example}

@@

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category
  which is not strong enforcing. The example is somehow
  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but it needs to be so, since natural examples are all in the well-beahved class mentioned above. 
  .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

A first class consists of those rewriting systems over
possibly hierarchical graphical structures obtained as algebras of
suitable signatures where rules are constrained not to merge elements
of top level sorts in the hierarchy (e.g., for graphs, nodes can be
merged while edges cannot). The idea here is to consider as graphs presheaves on categories in which there are object which plays the roles of \emph{roots}, i.e.~objects which are not codomains of any arrows beside the identity. \todo{D: quest'ultima frase ha senso?}%We first introduce signatures that are
%essentially acyclic graphs whose nodes can be seen as sorts.

\begin{definition}[Root-preserving graphical rewriting systems]
	Let $\X$ be a category, an object $X\in \X$ is a \emph{root} if the only arrow with codomain $X$ is $\id{X}$. The category The
	category $\gph{X}$ of \emph{$\X$-graphs} is the category
	$\Set^{\X}$. A
	\emph{root-preserving graphical rewriting system} is a left-linear
	rewriting system $(\gph{X}, \R)$ such that  the
	following hold for each rule
	$(l\colon K\to L, r\colon K\to R)$ in $\R$
	\begin{enumerate}
		\item for every $X\in \X$ and $x\in L(X)$, there exists a root $Y$
		 and an arrow $f\colon Y\to X$ such that $x$ belongs to the image of
		$L(f)\colon L(Y)\to L(X)$;
		\item $r\colon K\to R$ is mono on the roots, i.e.~for every root $X\in \X$, the component $r_X:K(X)\to R(X)$ is injective.
	\end{enumerate}
\end{definition}

\iffalse 
\begin{definition}
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph where $s,t\colon E\rightrightarrows V$.
  We define the \emph{depth} of a vertex $v\in V$ as\todo{ma ci serve tutta la depth?}
  \[
    \dph(v)=\begin{cases}
      0 & \text{there is no $e\in E$ such that $t(e)=v$}\\
      m+1 &\text{where } m=\min\{\dph(s(e)) \mid e\in E \text{ such that } t(e)=v
      \} 
    \end{cases}
  \]
\end{definition}

@@

\begin{definition}[root-preserving graphical rewriting systems]
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph
  and $\G$ the free category over $\mathcal{G}$. The
  category $\gph{G}$ of \emph{$\mathcal{G}$-graphs} is the category
  $\Set^{\G}$. A
  \emph{root-preserving graphical rewriting system}\todo{altro nome? source/leaf preserving} is a left-linear
  rewriting system $(\gph{G}, \R)$ such that  the
  following hold for each rule
  $(l\colon K\to L, r\colon K\to R)$ in $\R$
  \begin{enumerate}
  \item for every $v\in V$ and $x\in L(v)$, there exists a vertex $w$
    with depth $0$ and an arrow $p\colon w\to v$ (i.e.~a path in
    $\mathcal{G}$) such that $x$ belongs to the image of
    $L(p)\colon L(w)\to L(v)$;
  \item $r\colon K\to R$ is mono on vertices of depth $0$, i.e.~for every
    $v\in V$ such that $\dph(v)=0$, the component $r_v:K(v)\to R(v)$
    is mono.
  \end{enumerate}
\end{definition}
For instance, the category $\cat{Graph}$ can be obtained
by taking as $\X$ the category generated by $E \rightrightarrows V$. The sort
of edges $e$ has depth $0$ while $v$, the sort of nodes, has depth
$1$. Hence, condition $1$ asks that in the left-hand side of each rule
there are no isolated nodes, while condition $2$ asks that the
morphism $r: K \to R$ is mono on edges, i.e.~it can only merge nodes.
\fi

For instance, the category $\cat{Graph}$ can be obtained
by taking as $\X$ the category generated by $E \rightrightarrows V$. In this case $E$ is the only root, hence, condition $1$ asks that in the left-hand side of each rule
there are no isolated nodes, while condition $2$ asks that the
morphism $r: K \to R$ is mono on edges, i.e.~it can only merge nodes.

@@

\begin{restatable}{lemma}{lemVTame}
\label{bono}
  All root-preserving graphical rewriting systems are well-switching.
\end{restatable}

@@

  \begin{figure}[t]
    \input{Figure/rules-cas.tex}    
    \caption{A new rewriting system in $\cat{Graph}$.}
    \label{fi:rules-cas}
  \end{figure}

@@

\begin{remark}\label{rem:po}
	Notice that $\gph{X}$ is adhesive and that pushouts are computed component-wise, thus if $g\colon F\to G$ is the pushout of $r\colon K\to R$, then $g_X\colon F(X)\to G(X)$ is mono for every root $X$.
\end{remark}

@@

\begin{proof}
	Every root-preserving graphical rewriting system is strong enforcing because $\gph{G}$ is in the class $\mathbb{B}^+$ defined in \Cref{app:fill}.  Consider the following diagram 
	\[\xymatrix@R=25pt@C=30pt{L_0 \ar[d]_{m_0}&& K_0
		\ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
		\ar@<-.5ex>@/^.35cm/[drrr]|(.23)\hole|(.34)\hole_(.4){i_0} 	\ar@<.5ex>@/^.35cm/[drrr]^(.2){j_0}|(.31)\hole |(.4)\hole
		\ar[dr]|(.3)\hole_{h_0} && L_1 \ar@<.5ex>@/_.35cm/[dlll]^(.4){i_1} \ar@<-.5ex>@/_.35cm/[dlll]_(.2){j_1}
		\ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
		\ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
		D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
		\ar[rr]_{g_1}&& G_2}\]

	Since $f_1$ is mono then $i_0=j_0$. For every  root $X$ we have
	\[g_{0,X}\circ i_{1,X}=m_{1,X}=g_{0,X}\circ j_{1,X}\]
	which entails $i_{1,X}=j_{1,X}$. Let now $Y$ be another object, and take $y\in L_1(Y)$. By definition, we have a root $X$ and an arrow $f:X\to Y$ and $x\in L_{1}(X)$ such that $L(f)(y)=x$. By naturality we have
	\[i_{1,Y}(y)=i_{1,Y}(L(f)(x))=L(f)(i_{1, X}(x))=L(f)(j_{1, X}(x))=j_{1,X}(L(f)(x))=j_{1,Y}(y)\]
Thus $i_{1,Y}=j_{1,Y}$, as wanted.
\end{proof}
